---
title: "Word embeddings"
output: html_output
---

```{r}
library(keras)
```

# Loading the IMDB data for use with an embedding layer

Restrict the movie reviews to the top 10,000 most common words then cut off the reviews after only 20 words to make the situation a bit simpler.

```{r}
# Number of most common words to consider as features
max_features <- 10000

# Loads the data as lists of integers
c(c(x_train, y_train), c(x_test, y_test)) %<-% dataset_imdb(num_words = max_features)

# Cut off the text after 20 words (i.e. among the max_features most common words)
maxlen <- 20

# Turns the lists of integers into a 2D integer tensor of shape (samples, maxlen)
x_train <- pad_sequences(x_train, maxlen = maxlen)
x_test <- pad_sequences(x_test, maxlen = maxlen)
```

```{r}
dim(x_test)
```

# Using an embedding layer and classifier on the IMDB data

The network will:

1. Learn 8-dimensional embeddings for each of the 10,000 words, 
2. Turn the input integer sequences (2D integer tensor) into embedded sequences (3D float tensor),
3. Flatten the tensor to 2D, and
4. Train a single dense layer on top for classification.

```{r}
# Part 1 & 2:
# Specify the maximum input length to the embedding layer so you can later flatten the embedded inputs.
model <- keras_model_sequential() %>%
  layer_embedding(input_dim = 10000, 
                  output_dim = 8,
                  input_length = maxlen)

# Part 3:
# After the embedding layer, the activations have shape (samples, maxlen, 8).
# Flattens the 3D tensor of embeddings into a 2D tensor of shape (samples, maxlen * 8)
model <- model %>% 
  layer_flatten()

# Part 4:
# Adds the classifier on top
model <- model %>%
  layer_dense(units = 1, activation = "sigmoid")

```

Compile the model

```{r}
model %>% compile(
  optimizer = "rmsprop",
  loss = "binary_crossentropy",
  metrics = c("acc")
)

summary(model)
```


Train the model

```{r}
history <- model %>% fit(
  x_train, y_train,
  epochs = 10,
  batch_size = 32,
  validation_split = 0.2
)
```


```{r}
plot(history)
```


```{r}
# Peak validation accuracy:
history$metrics$val_acc[10]
```

